Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-14 doc/thesis/intro.tex
Editing and proofreading intro

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
correspond to their common ancestors, {\color{blue}\uline{usually extinct
(although occasionally the internal nodes may be observed as well, eg.
\autocite{jombart2011reconstructing}).}} The edges or \defn{branches} of the
phylogeny connect ancestors to their descendants. Phylogenies may have a
\defn{root}, which is a node with no descendants distinguished as the most
recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

@@

The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
lengths connecting node $n$ to its $\nc(n)$ children. {\color{blue}\uline{
Furthermore, let $\nl(n)$ be the number of children of $n$ which are leaves (we
always have $\nl(n) \leq \nc(n)$). The \defn{production rule} of $n$ is the
pair $(\nc(n), \nl(n))$.}} That is, if two nodes have the same number of
children and among these, the same number of leaves, then they have the same
production rule. Let $k_G(x, y)$ be a Gaussian radial basis function of the
vectors $x$ and $y$,
\[
  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
\]
where $\norm{\cdot}_2$ is the Euclidean norm and $\sigma$ is a variance
parameter. The tree kernel is defined as 
\begin{align}
  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum_{n_2 \in N(T_2)} \Delta (n_1, n_2),
  \label{eq:kernel}
\end{align}
where
\[
  \Delta(n_1, n_2) =
  \begin{cases}
    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
    0 & \text{otherwise}.
  \end{cases}
\]
{\color{blue}\uline{
The parameter $\lambda$ in the above expression, is called the \emph{decay
factor}~\autocite{moschitti2006making}, and takes a value between 0 and 1.
Without this parameter, terms in the sum} \ref{eq:kernel} \uline{corresponding
to large subset trees with the same topology would be similarly large and tend
to dominate the kernel score. $\lambda$ penalizes $\Delta$ more strongly as the
number of recursive calls increases, which downweights the largest matching
substructures and allows smaller matches to contribute more to the kernel
score. }} In this work, we refer to the parameters $\lambda$ and $\sigma$ as
\defn{meta-parameters}, to avoid confusing them with model parameters we are
trying to estimate. {\color{blue}\uline{ When evaluating the tree kernel, it is
helpful to reorder the children of each internal node such that the larger of
the two subtrees is on the right-hand side. If the two subtrees have equal
sizes, then the child with the longer branch length can be put on the
right-hand side. This operation is referred to as \defn{ladderizing}. Since the
ordering of children is arbitrary in phylogenies, this operation ensures that a
maximal number of matching subset trees are counted by the tree kernel without 
making meaningful changes to the trees. }}

{\color{blue}\uline{
The tree kernel was later shown to be highly effective in differentiating trees
simulated under a compartmental model with two risk groups of varying contact
rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC}} (see \cref{sec:abc}), \uline{to fit epidemiological models to
observed trees. }}

@@

Several properties of scale-free networks are relevant in epidemiology.  The
high-degree hub nodes are known as
\defn{superspreaders}~\autocite{kemper1980identification}, which have been
postulated to contribute in varying degree to the spread of diseases such as
\gls{HIV}~\autocite{stadler2013uncovering} and
\gls{SARS}~\autocite{shen2004superspreading}. Scale-free networks have no
epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
arbitrarily low transmissibility {\color{red}\sout{can persist}} 
{\color{blue}\uline{have a chance, however small, of persisting}} at low levels
indefinitely. This is in contrast with homogeneously mixed populations, in
which transmissibility below the epidemic threshold would result in exponential
decay in the number of infected individuals and eventual extinction of the
pathogen~\autocite{anderson1992infectious}.

@@

There has been some contention over the idea that contact networks are
scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
partner formation to empirical degree distributions derived from population
surveys of sexual behaviour. They found that a negative binomial distribution,
rather than a power law, was the best fit to five out of six datasets, although
the difference in goodness of fit was extremely small in four out of these
five. \textcite{bansal2007individual} found that an exponential distribution,
rather than a power law, was the best fit to degree distributions of six social
and sexual networks. \textcite{drombowski2013} contend that sexual networks are
shaped more by homophily (``like attract like'') than by preferential
attachment, but find that \gls{IDU} network do demonstrate a scale-free
structure.

In the paper describing the \gls{BA} model, \citeauthor{barabasi1999emergence}
suggest an extension where the probability of choosing a partner of degree $d$
is proportional to $d^\alpha + 1$ for some constant \gls{alpha}.
{\color{blue}\uline{When $\gls{alpha} \neq 1$, the degree distribution no
longer follows a power law~\autocite{krapivsky2000connectivity}. For
$\gls{alpha} < 1$, the distribution is a stretched exponential, meaning that
the number of nodes of degree $k$ is proportional to $\exp(-k^\beta)$ for some
constant $\beta$. For $\gls{alpha} > 1$, the distribution takes on a
characteristic called \emph{gelation}, where a one or a few high-degree hub
nodes are connected to nearly every other node in the graph. We do not believe
these departures from the power law affect the applicability of the model to
real world networks. In fact,} \textcite{de2007preferential} \uline{were able to
estimate the \acrlong{PA} power from partner count data collected from the same
individuals for consecutive time intervals, and found a value less than one in
all cases. It is also worth noting that, in addition to the \gls{BA} model,
other investigations of the interaction between contact networks and
transmission trees have studied the \acrlong{ER} and \acrlong{WS}
models~\autocite{leventhal2012inferring}, whose degree distributions do not
generally follow a power law under any parameter settings.}}

@@

{\color{blue}\uline{
Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. In particular, for a network model with parameters $\theta$ and an input
transmission tree $T$, we are interested in the posterior distribution}
\begin{align}
  \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
  \label{eq:post}
\end{align}
\uline{where we have adopted the common abuse of the symbol $\Pr$ to refer to a
probability density. As alluded to in} \cref{sec:obj}, \uline{both the
likelihood $\Pr(T \mid \theta)$ and the normalizing constant $\Pr(T)$ are
likely computationally intractable (this will be explained further in}
\cref{subsec:just}\uline{). Hence, rather than computing the posterior
distribution analytically, we will approximate it using a \defn{Monte Carlo}
approach. The fundamental idea behind Monte Carlo methods is succinctly
expressed by \textcite{liu2001theoretical}}:
\begin{quote}
    \uline{Monte Carlo's view of the world is that any probability distribution $\pi$, 
    regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any computation
    of expectations using $\pi$ can be replaced to an acceptable degree of
    accuracy by using the empirical distribution resulting from the discrete
    sample.}
\end{quote}
\uline{In other words, if we are able to sample enough points from a
distribution of interest, we will be able to make reasonably accurate
statements about the distribution itself. For example, the expected value of
the distribution can be estimated by the sample's population mean. The reason
Monte Carlo methods will be useful in this work is that algorithms exist for
obtaining samples from distributions that are analytically intractable and from
which direct sampling is not possible } (for a review see
\autocite{robert2004monte}). \uline{\Gls{SMC}}~\autocite{doucet2000sequential,
doucet2001introduction, liu2008monte} \uline{is one such algorithm.}

\uline{\Gls{SMC} is also known as the \emph{particle filter}. Rather than
sampling points one at a time from the target distribution, \gls{SMC} considers
a population of points or ``particles'', here denoted $\set{x^{(k)}}$ and
indexed by an integer $k$. The particles are associated with weights,
$\set{w(x^{(k)})}$. These weighted particles are a representation for the
target distribution. For example, the expected value of the distribution is
approximated by}
\[
  \frac{1}{n} \sum_{k=1}^n x^{(k)} w(x^{(k)}),
\]
\uline{where $n$ is the number of particles. Initially, the particles do not
represent the target distribution but rather a more tractable distribution from
which direct sampling is straightforward. The word ``sequential'' is used to
describe the iterative process of perturbation, resampling, and reweighting
applied to the particles in such a way that they converge, collectively, to a
sample from the target.}

\uline{In this work, the distribution of interest is the posterior distribution}
\ref{eq:post}. \uline{The particles are particular values of the parameters
$\theta$ of the contact network model being studied. If we were taking a
typical Bayesian Monte Carlo approach to this problem, the particles would end
up weighted by their posterior probability and distributed in such a way that
the weighted population was a reasonable representation of $\Pr(\theta \mid
T)$. In our case, due to the intractable likelihood, we will need to consider
an approximation to the posterior } (see \cref{sec:abc}). \uline{However, for
now, nothing is lost by assuming that our target distribution is the posterior
itself.}

\uline{The goal of this section of the introduction is to describe an algorithm
called the \gls{SMC} sampler~\autocite{del2006sequential}, which forms the
basis of the adaptive \gls{ABC}-\gls{SMC} algorithm we apply toward the main
objective of this work. We begin by describing \gls{SIS}, which is a precursor
to \gls{SMC} that samples from a sequence of distributions defined on spaces of
increasing dimension. We then describe \gls{SMC} itself, which extends
\gls{SIS} with a resampling step to fight particle degeneracy. Finally, we
outline the \gls{SMC} sampler, which allows \gls{SMC} to be applied to
sequences of distributions all defined on the same space. This terminology will
become clear as the methods are described. }}

{\color{red}\sout{Sequential Monte Carlo (SMC) is the name for a family of
statistical inference methods that rely on approximating probability
distributions of interest with large collections of \defn{particles}, here
denoted $\set{x^{(k)}}$}\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte}. \sout{These collections or \defn{populations} are constructed to form
a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
meaning that the empirical distribution of the particles converges in
distribution to $\pi$ as the population size gets
large}\autocite{liu2001theoretical}. \sout{The word \defn{sequential} is used because
the particle populations are modified in an iterative fashion over time, for
example, to incorporate new evidence.}}

To fully describe \gls{SMC}, we will introduce some notation and terminology.
The definitions of these terms will become clearer as they are used. For a
sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the partial
sequence $x_1, \ldots, x_i$. The subscript $^{(k)}$ will be used to indicate
the $k$th particle in a population. To ease the notational burden we will omit
the superscripts and subscripts on the weight functions $w$.

We define a \defn{Markov kernel} as the continuous analogue of the transition
matrix in a finite-state Markov model.  For some spaces $X$ and $Y$, $K \maps X
\times Y \to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more rigorous definition can be found in
\eg~\autocite{kallenberg2006foundations}. Note that Markov kernels have nothing
to do with the kernel functions defined in \cref{subsec:treeshape}, other than
sharing a name (the word ``kernel'' is ubiquitous in mathematics).

@@

\Gls{SIS}~\autocite{gordon1993novel} {\color{blue}\uline{is a particle-based
method}} whose aim is to sample from a distribution $\pi$ on an
high-dimensional space, say $\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis
of \gls{SIS} is \gls{IS}, which is a method of estimating summary statistics of
distributions which are known only up to a normalizing constant, and therefore
cannot be sampled from directly. That is, if $\pi$ is such a distribution and
$f$ is any real-valued function, \gls{IS} is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can write
the expectation of interest as
{\color{blue}
\begin{align}
  \int f(x) \pi(x) \d x = \frac{1}{Z} \int w(x) f(x) \d x.
  \label{eq:is}
\end{align}
}
{\color{blue}\uline{
Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can both be
evaluated pointwise, the integral $\int w(x) f(x) \d x$ can be approximated by
a Monte Carlo estimate. Moreover, the normalizing constant $Z$ can be expressed
in terms of the importance weight and distribution, $Z = \int w(x) \eta(x) \d
x$. Therefore, we have all the ingredients we need to obtain an estimate of
$\pi(f)$ using} \cref{eq:is}.} Although this is a simple and elegant approach,
the drawback is that the variance of the estimate is proportional to the
variance of the importance weights~\autocite{liu2008monte}, which may be quite
large if $\eta$ and $\gamma$ are very different. Therefore, the practical use
of \gls{IS} on its own is limited, since it depends on finding an importance
distribution similar to $\pi$, which we usually know very little about
\textit{a priori}.

The objective of \gls{SIS} is to build up an importance distribution $\eta$ for
$\pi$ sequentially. By the general product rule, $\pi(\vec{x})$ can be
decomposed as
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ can be written recursively as
\begin{align}
  \label{eq:sisw}
  w(\vec{x_i}) = \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
  = \frac{\pi(x_i \mid \vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}}).
\end{align}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. This choice is made on a
problem-specific basis, taking any available information about $\pi(x_i \mid
\vec{x_{i-1}})$ into account (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).  One potential
choice for $\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid
\vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
prior distribution may be used. The exact form of $\eta(x_i \mid
\vec{x_{i-1}})$ which minimizes the variance of the weights is called the
\defn{optimal kernel}~\autocite{cappe2007overview}, the name deriving from the
fact that $k(x_i, \vec{x_{i-1}}) = \eta(x_i \mid \vec{x_{i-1}})$ is a Markov
kernel. In some applications, it is possible to approximate the optimal kernel
or even compute it explicitly.

The recursive definition \ref{eq:sisw} suggests an algorithm for obtaining a
sample from {\color{blue}\uline{$\eta$ and using it to obtain an approximate
sample from $\pi$ by \gls{IS}}} (\cref{alg:sis}).  We begin with $n$
``particles'', which have been sampled from the importance distribution
$\eta(x_0)$ for $\pi(x_0)$. The particles are updated and reweighted $d$ times,
corresponding to the $d$ elements of the decomposition of $\pi$. At the $i$th
step, each particle is extended to include $x_i$ drawn according to the chosen
$\eta(x_i \mid \vec{x_{i-1}})$, and the importance weights are recalculated and
normalized.

>>>>

In evolutionary biology, a \defn{phylogeny}, or \defn{phylogenetic tree}, is a
graphical representation of the evolutionary relationships among a group of
organisms or species (generally, \defn{taxa})~\autocite{haeckel1866generelle}.
The \defn{tips} of a phylogeny, that is, the nodes without any descendants,
correspond to \defn{extant}, or observed, taxa. The \defn{internal nodes}
correspond to their common ancestors, \add{usually extinct (although
occasionally the internal nodes may be observed as well, eg.
\autocite{jombart2011reconstructing}).} The edges or \defn{branches} of the
phylogeny connect ancestors to their descendants. Phylogenies may have a
\defn{root}, which is a node with no descendants distinguished as the most
recent common ancestor of all the extant
taxa~\autocite{harding1971probabilities}. When such a root exists, the tree is
referred to as being \defn{rooted}; otherwise, it is \defn{unrooted}. The
structural arrangement of nodes and edges in the tree is referred to as its
\defn{topology}~\autocite{cavalli1967phylogenetic}.

@@

The kernel score of a pair of trees, denoted $K(T_1, T_2)$, is defined as a sum
over all pairs of nodes $(n_1, n_2)$, where $n_1$ is a node in $T_1$ and $n_2$
is a node in $T_2$. Following \textcite{poon2013mapping}, let $N(T)$ denote the
set of all nodes in $T$, $\nc(n)$ be the number of children of node $n$,
$c_n^{j}$ be the $j$th child of node $n$, and $l_n$ be the vector of branch
lengths connecting node $n$ to its $\nc(n)$ children. \add{Furthermore, let
$\nl(n)$ be the number of children of $n$ which are leaves (we always have
$\nl(n) \leq \nc(n)$). The \defn{production rule} of $n$ is the pair $(\nc(n),
\nl(n))$.} That is, if two nodes have the same number of children and among
these, the same number of leaves, then they have the same production rule. Let
$k_G(x, y)$ be a Gaussian radial basis function of the vectors $x$ and $y$,
\[
  k_G(x, y) = \exp\left(-\frac{1}{2\sigma} \norm{x - y}_2^2\right),
\]
where $\norm{\cdot}_2$ is the Euclidean norm and $\sigma$ is a variance
parameter. The tree kernel is defined as 
\begin{align}
  K(T_1, T_2) = \sum_{n_1 \in N(T_1)} \sum_{n_2 \in N(T_2)} \Delta (n_1, n_2),
  \label{eq:kernel}
\end{align}
where
\[
  \Delta(n_1, n_2) =
  \begin{cases}
    \lambda & n_1 \text{ and } n_2 \text{ are leaves} \\
    \lambda k_G(l_{n_1}, l_{n_2}) \displaystyle\prod_{j=1}^{\nc(n_1)} \left(1 +
    \Delta(c_{n_1}^j, c_{n_2}^j) \right) & \begin{aligned} n_1 \text{ and } n_2 \text{ have the same} \\ \text{production rule} \end{aligned} \\
    0 & \text{otherwise}.
  \end{cases}
\]
\add{The parameter $\lambda$ in the above expression, is called the \emph{decay
factor}~\autocite{moschitti2006making}, and takes a value between 0 and 1.
Without this parameter, terms in the sum \ref{eq:kernel} corresponding to large
subset trees with the same topology would be similarly large and tend to
dominate the kernel score. $\lambda$ penalizes $\Delta$ more strongly as the
number of recursive calls increases, which downweights the largest matching
substructures and allows smaller matches to contribute more to the kernel
score.} In this work, we refer to the parameters $\lambda$ and $\sigma$ as
\defn{meta-parameters}, to avoid confusing them with model parameters we are
trying to estimate. \add{When evaluating the tree kernel, it is helpful to
reorder the children of each internal node such that the larger of the two
subtrees is on the right-hand side. If the two subtrees have equal sizes, then
the child with the longer branch length can be put on the right-hand side. This
operation is referred to as \defn{ladderizing}. Since the ordering of children
is arbitrary in phylogenies, this operation ensures that a maximal number of
matching subset trees are counted by the tree kernel without making meaningful
changes to the trees.}

\add{The tree kernel was later shown to be highly effective in differentiating
trees simulated under a compartmental model with two risk groups of varying
contact rates~\autocite{poon2015phylodynamic}. In that paper,
\citeauthor{poon2015phylodynamic} used the tree kernel as the distance function
in \gls{ABC} (see \cref{sec:abc}), to fit epidemiological models to observed
trees.}

@@

Several properties of scale-free networks are relevant in epidemiology.  The
high-degree hub nodes are known as
\defn{superspreaders}~\autocite{kemper1980identification}, which have been
postulated to contribute in varying degree to the spread of diseases such as
\gls{HIV}~\autocite{stadler2013uncovering} and
\gls{SARS}~\autocite{shen2004superspreading}. Scale-free networks have no
epidemic threshold~\autocite{pastor2001epidemic}, meaning that diseases with
arbitrarily low transmissibility \del{can persist} \add{have a chance, however
small, of persisting} at low levels indefinitely. This is in contrast with
homogeneously mixed populations, in which transmissibility below the epidemic
threshold would result in exponential decay in the number of infected
individuals and eventual extinction of the
pathogen~\autocite{anderson1992infectious}.

@@

There has been some contention over the idea that contact networks are
scale-free. \textcite{handcock2004likelihood} fit several stochastic models of
partner formation to empirical degree distributions derived from population
surveys of sexual behaviour. They found that a negative binomial distribution,
rather than a power law, was the best fit to five out of six datasets, although
the difference in goodness of fit was extremely small in four out of these
five. \textcite{bansal2007individual} found that an exponential distribution,
rather than a power law, was the best fit to degree distributions of six social
and sexual networks. \add{\textcite{dombrowski2013topological} contend that
sexual networks are shaped more by homophily (``like attract like'') than by
preferential attachment, but find that \gls{IDU} network do demonstrate a
scale-free structure.}

In the paper describing the \gls{BA} model, \citeauthor{barabasi1999emergence}
suggest an extension where the probability of choosing a partner of degree $d$
is proportional to $d^\alpha + 1$ for some constant \gls{alpha}.
\add{When $\gls{alpha} \neq 1$, the degree distribution no longer follows
a power law~\autocite{krapivsky2000connectivity}. For $\gls{alpha} < 1$, the
distribution is a stretched exponential, meaning that the number of nodes of
degree $k$ is proportional to $\exp(-k^\beta)$ for some constant $\beta$. For
$\gls{alpha} > 1$, the distribution takes on a characteristic called
\emph{gelation}, where a one or a few high-degree hub nodes are connected to
nearly every other node in the graph. We do not believe these departures from
the power law affect the applicability of the model to real world networks. In
fact, \textcite{de2007preferential} were able to estimate the \acrlong{PA}
power from partner count data collected from the same individuals for
consecutive time intervals, and found a value less than one in all cases. It is
also worth noting that, in addition to the \gls{BA} model, other investigations
of the interaction between contact networks and transmission trees have studied
the \acrlong{ER} and \acrlong{WS} models~\autocite{leventhal2012inferring},
whose degree distributions do not generally follow a power law under any
parameter settings.}

@@

\add{Recall that the primary objective of our work is to develop a statistical
inference method for estimating contact network parameters from transmission
trees. For a network model with parameters $\theta$ and an input transmission
tree $T$, our goal is to estimate statistics, such as means and credible
intervals, of the posterior distribution}
\begin{align}
  \madd{\post(\theta \mid T) = \frac{\lik(T \mid \theta) \prior(\theta)}
    {\int_\Theta\lik(T \mid \theta) \prior(\theta)}.}
  \label{eq:post}
\end{align}
\add{As alluded to in \cref{sec:obj}, both the likelihood $\lik(T \mid \theta)$
and the normalizing constant are likely computationally intractable
(this will be discussed further in \cref{subsec:just}). Hence, rather than
computing the posterior distribution analytically, we will approximate it using
a \defn{Monte Carlo} approach. The fundamental idea behind Monte Carlo methods
is succinctly expressed by \textcite{liu2001theoretical}:}
\begin{quote}
    \add{Monte Carlo's view of the world is that any probability distribution
    $\pi$, regardless of its complexity, can always be \emph{represented} by a
    discrete sample from it. By ``represented'', we mean that any
    computation of expectations using $\pi$ can be replaced to an acceptable
    degree of accuracy by using the empirical distribution resulting from the
    discrete sample.}
\end{quote}
\add{In other words, if we are able to sample enough points from a distribution
of interest, we will be able to make reasonably accurate statements about the
distribution itself. For example, the expected value of the distribution can be
estimated by the sample's population mean. The reason Monte Carlo methods will
be useful in this work is that algorithms exist for obtaining samples from
distributions that are analytically intractable and from which direct sampling
is not possible (for a review see \autocite{robert2004monte}).
\Gls{SMC}~\autocite{doucet2000sequential, doucet2001introduction, liu2008monte}
is one such algorithm.}

\add{\Gls{SMC} considers a population of points or ``particles'', here denoted
$\set{x^{(k)}}$ and indexed by an integer $k$. The particles are associated
with weights, $\set{w(x^{(k)})}$. After the algorithm has been run, the
weighted particles are a Monte Carlo representation for the target
distribution. For example, the expected value is approximated by}
\[
  \madd{\frac{1}{n} \sum_{k=1}^n x^{(k)} w(x^{(k)}),}
\]
\add{where $n$ is the number of particles. Initially, the particles do not
represent the target distribution but rather a more tractable distribution from
which direct sampling is straightforward. The word ``sequential'' is used to
describe the iterative process of perturbation, resampling, and reweighting
applied to the particles in such a way that they converge, collectively, to a
representation of the target. \Gls{SMC} is also known as the \defn{particle
filter}.}

\add{In this work, the distribution of interest is the posterior distribution
\ref{eq:post}. The particles are particular values of the parameters $\theta$
of the contact network model being studied. If we were taking a typical
Bayesian Monte Carlo approach to this problem, the particles would end up
weighted by their posterior probability and distributed in such a way that the
weighted population was a reasonable representation of $\post(\theta \mid T)$.
In our case, due to the intractable likelihood, we will need to consider an
approximation to the posterior (see \cref{sec:abc}). However, for now, nothing
is lost by assuming that our target distribution is the posterior itself.}

\add{Here we describe an algorithm called the \gls{SMC} sampler, developed by
\textcite{del2006sequential}, which forms the basis of the adaptive
\gls{ABC}-\gls{SMC} algorithm we apply toward the main objective of this work.
We begin by describing \gls{SIS}, which is a precursor to \gls{SMC} that
samples from a sequence of distributions defined on spaces of increasing
dimension. We then describe \gls{SMC} itself, which extends \gls{SIS} with a
resampling step to fight particle degeneracy. Finally, we outline the \gls{SMC}
sampler, which allows \gls{SMC} to be applied to sequences of distributions all
defined on the same space. This terminology will become clear as the methods
are described.}

\del{\Gls{SMC} is the name for a family of statistical inference methods that
rely on approximating probability distributions of interest with large
collections of \defn{particles}, here denoted
$\set{x^{(k)}}$~\autocite{doucet2000sequential, doucet2001introduction,
liu2008monte}. These collections or \defn{populations} are constructed to form
a \defn{Monte Carlo approximation} to some distribution of interest $\pi$,
meaning that the empirical distribution of the particles converges in
distribution to $\pi$ as the population size gets
large~\autocite{liu2001theoretical}. The word \defn{sequential} is used because
the particle populations are modified in an iterative fashion over time, for
example, to incorporate new evidence.}

To fully describe \gls{SMC}, we will introduce some notation and terminology.
The definitions of these terms will become clearer as they are used. For a
sequence $x_1, \ldots, x_d$, we will write $\vec{x_i}$ to mean the partial
sequence $x_1, \ldots, x_i$. The subscript $^{(k)}$ will be used to indicate
the $k$th particle in a population. To ease the notational burden we will omit
the superscripts and subscripts on the weight functions $w$. We define a
\defn{Markov kernel} as the continuous analogue of the transition matrix in a
finite-state Markov model. For some spaces $X$ and $Y$, $K \maps X \times Y
\to [0, 1]$ such that
\begin{align}
    \label{eq:mk}
    \int_Y K(x, y) \d y = 1
\end{align}
for all $x \in X$. This is an ``operational'' definition of Markov kernel which
will be suitable for our purposes. A more rigorous definition can be found in
\eg~\autocite{kallenberg2006foundations}. Note that Markov kernels have nothing
to do with the kernel functions defined in \cref{subsec:treeshape}, other than
sharing a name (the word ``kernel'' is ubiquitous in mathematics). \add{Also, 
the variable $\pi$ is used here for a generic target distribution to describe
the algorithm, and does not refer to the prior or posterior distributions we
will eventually want to work with.}

@@

\Gls{SIS}~\autocite{gordon1993novel} \add{is a particle-based method} whose aim
is to sample from a distribution $\pi$ on an high-dimensional space, say
$\pi(\vec{x}) = \pi(x_1, \ldots, x_d)$. The basis of \gls{SIS} is \gls{IS},
which is a method of estimating summary statistics of distributions which are
known only up to a normalizing constant, and therefore cannot be sampled from
directly. That is, if $\pi$ is such a distribution and $f$ is any real-valued
function, \gls{IS} is concerned with estimating
\[
  \pi(f) = \int f(x)\pi(x)\d x = \int f(x) \frac{\gamma(x)}{Z} \d x,
\]
where the integral is over the space on which $\pi$ is defined, $\gamma(x)$ is
known pointwise, and $Z = \int \gamma(x) \d x$ is the unknown normalizing
constant. Suppose we have at hand another distribution $\eta$, called the
\defn{importance distribution}, from which we are able to sample. Define the
\defn{importance weight} as the ratio $w(x) = \gamma(x)/\eta(x)$. We can write
the expectation of interest as
\begin{align}
  \int f(x) \pi(x) \d x = 
  \mdel{\dfrac{\int f(x) \gamma(x) \d x}{\int w(x) \eta(x) \d x}}\quad
  \madd{\frac{1}{Z} \int w(x) \eta(x) f(x) \d x}.
  \label{eq:is}
\end{align}
\add{Since $\eta$ can be sampled from exactly, and $\gamma$ and $f$ can both be
evaluated pointwise, the integral $\int w(x) \eta(x) f(x) \d x$ can be
approximated by a Monte Carlo estimate. We simply take a sample from $\eta$,
multiply each point in the sample by $f$ and $\gamma$ evaluated at that point,
and sum the results. Moreover, the normalizing constant $Z$ can be expressed in
terms of the importance weight and distribution, $Z = \int w(x) \eta(x) \d x$.
Therefore, we have all the ingredients we need to obtain an estimate of
$\pi(f)$ using \cref{eq:is}.} Although this is a simple and elegant approach,
the drawback is that the variance of the estimate is proportional to the
variance of the importance weights~\autocite{liu2008monte}, which may be quite
large if $\eta$ and $\gamma$ are very different. Therefore, the practical use
of \gls{IS} on its own is limited, since it depends on finding an importance
distribution similar to $\pi$, which we usually know very little about
\textit{a priori}.

The objective of \gls{SIS} is to build up an importance distribution $\eta$ for
$\pi$ sequentially. By the general product rule, $\pi(\vec{x})$ can be
decomposed as
\[
  \pi(\vec{x}) 
  = \pi(x_1) \pi(x_2 \mid x_1) \cdots
    \pi(x_{d-1} \mid \vec{x_{d-2}}) \pi(x_d \mid \vec{x_{d-1}}).
\]
This decomposition is natural in many contexts, particularly for on-line
estimation. For example, in a stateful model like an \gls{HMM}, $x_i$ may
represent the state at time $i$, with $\pi(\vec{x})$ being the posterior
distribution over possible paths. The importance distribution $\eta$ for $\pi$
will be constructed using a similar decomposition,
\[
  \eta(\vec{x}) 
  = \eta(x_1) \eta(x_2 \mid x_1) \cdots
    \eta(x_{d-1} \mid \vec{x_{d-2}}) \eta(x_d \mid \vec{x_{d-1}}).
\]
The importance weights for $\eta$ can be written recursively as
\begin{equation}
  \label{eq:sisw}
  \begin{alignedat}{3}
  w(\vec{x_i}) 
    &= \frac{\pi(\vec{x_i})}{\eta(\vec{x_i})}
     & \text{definition of importance weight} \\
    &= \frac{\pi(x_i \mid \vec{x_{i-1}})\pi(\vec{x_{i-1}})}
         {\eta(x_i \mid \vec{x_{i-1}})\eta(\vec{x_{i-1}})}
      & \qquad\qquad\text{definition of conditional probability} \\
    &= \frac{\pi(x_i \mid \vec{x_{i-1}})}
            {\eta(x_i \mid \vec{x_{i-1}})}\cdot w(\vec{x_{i-1}})
      & \text{definition of importance weight}.
  \end{alignedat}
\end{equation}
Thus, we can choose $\eta(x_i \mid \vec{x_{i-1}})$ such that the variance of
the importance weights is as small as possible at every step, eventually
arriving at a full importance distribution. This choice is made on a
problem-specific basis, taking any available information about $\pi(x_i \mid
\vec{x_{i-1}})$ into account (see \textit{e.g.}
\autocite{smith2013sequential,liu2008monte} for many examples).  One potential
choice for $\eta(x_i \mid \vec{x_{i-1}})$ is simply $\pi(x_i \mid
\vec{x_{i-1}})$, if it is possible to compute. In a Bayesian setting, the
prior distribution may be used. The exact form of $\eta(x_i \mid
\vec{x_{i-1}})$ which minimizes the variance of the weights is called the
\defn{optimal kernel}~\autocite{cappe2007overview}, the name deriving from the
fact that $k(x_i, \vec{x_{i-1}}) = \eta(x_i \mid \vec{x_{i-1}})$ is a Markov
kernel. In some applications, it is possible to approximate the optimal kernel
or even compute it explicitly.
